---
title: "Analysis Report 2: Influences of Non-small Cell Lung Cancer on Lipid Metabolism"
author: "Allison Bogisich"
date: "November 15, 2017"
output: github_document
bibliography: references.bib
csl: bioinformatics.csl
---

*Overall, a single-spaced page is about 500 words. So if the guidelines say half of a page, think about writing around 250 words. You can use the wordcountaddin in RStudio to track your progress.*

# Introduction

One of the most longstanding symptoms of cancer has been increased rates of lipid synthesis, as
cancer cells are prone to exhibit specific alterations in their metabolic activities. Cancer is
predicted to kill approximately 595,690 Americans in 2016, equating to about 1,630 people per
day according to the American Cancer Society. Cancer-associated metabolic reprogramming
occurs in order to support the increased production of cells via necessary increased rates of
biosynthesis of proteins, nucleic acids, and aforementioned lipids [@baenke2013hooked]. However,
not until relatively recently has research delving into the metabolic processes involved in lipid
biosynthesis and their regulation- within the context of human disease- sustained much traction.
The unique connection between lipid metabolism and cancer in humans lies within its linkage to
epidermal growth factor (EGF) proteins and their receptors [@normanno2006epidermal]. Non-small
cell lung cancer for example, is EGF specific and characterized by dramatic changes specifically
in phospholipid profiles [@marien2015non]. In 2017, non-small cell lung cancer is the most prevalent type of lung cancer, accounting for about 85 % of all lung cancer cases in the USA according to the American Cancer Society.There are many human lipid metabolism disorders associated with aberrant lipid homeostasis besides cancers, including but not limited to:Gaucher’s disease, Faber’s disease, Wolman’s disease and many others [@saltiel2001insulin, @madhusudan2004multicenter,@chanarin1975neutral]. While these diseases are fairly rare, there are few current options for treatment and the survival rates-with decent quality of life-are quite low.


SNORAs and SNORDs are small noncoding RNAs involved in RNA processing and nogalamycin biosynthesis. This biosynthesis of noglamycin is critical to antitumor and antibiotic functionality of cellular maintenance. Both types have been found to be dysregulated in various types of cancer and have been implicated in not only the development but also the progressions of human cancer malignancy. This additional analysis will be based on data derivied from the previous work of Li and Seo et al. in non-small cell cancer patients. I hypothesize that the expression of SNOR genes in the patient tissues examined by Li and Seo et al. will show a markeded flux in their levels of expression. That in the patients will later stage cancers, there will be more expression relative to the earlier stage cancer patients. I also predict that the smoking patients will have a higher degree of abberant expression than that of the non-smokers, due to the additional carcinogens found in tobacco products [@freour2008active].

# Methods

## Sample origin and sequencing

The work of Li and her team [@li2015rna] included data from the first large scale RNA sequencing study of lung adenocarcinoma by Seo et al. [@seo2012transcriptional]. Paired surgical samples of both cancerous lung tissue and non-cancerous tissue were taken from 200 Korean patients. Only eighty-seven samples were then sequenced using RNA-seq and the Illumina TruSeq SBS Kit v3 on a HiSeq 2000 sequencer (Illumina Inc.) to obtain 100-bp paired-end reads. Of the 87 individuals whose cancer specimens were screened and sequenced, smoking history prior to diagnosis of lung cancer was provided by only 83 (47 smokers, 36 never-smokers, and four of unknown history).They were then aligned and read based on the NCBI human reference genome assembly build (build 37.1). Li et al. then downloaded Seo et al.'s RNA sequencing and phenotypic metadata for 68 patients(with 34 validated smoking and 34 non-smoking histories) from Gene Expression Omnibus (GEO) with accession number GSE40419. Comparisons were conducted based on phenotypes, including age at diagnosis,gender, and smoking history which are available from the public website (http://genome.cshlp.org/content/22/11/2109/suppl/DC1). Additionally, in order to validate genes identified in the nonsmoker group, independent RNA-seq data from six non-smoker patients were also downloaded from another GEO with accessionnumber GSE37765 [@kim2013high]. Differential gene expression analysis was then conducted between paired normal and tumor tissue using R Bioconductor and edgeR. Genotype calling was done for single-nucleotide variants, from RNA-seq bam files, with pair-end RNA-seq reads aligned to humangenome assembly Ensembl GRCh37 by Tophat.

## Computational

These are the methods that were used to process the sequencing data. Should probably be at least a half of a page. At a very minimum should include citations for biomartr, trimmomatic, and sailfish. Note that these three methods references don't count towards the five references you need to cite in the introduction.

# Results

In addition to a minimum of 4-5 figures/tables (and associated captions), you should include sufficient text in this section to describe what your findings were. Remember that in the results section you just describe what you found, but you don't interpret it - that happens in the discussion.

```{r load-libraries, message = FALSE, echo = FALSE}
# Be sure to install these packages before running this script
# They can be installed either with the intall.packages() function
# or with the 'Packages' pane in RStudio

# load general-use packages
library("dplyr")
library("tidyr")
library("knitr")
library("ggplot2")
library("magrittr")

# this package allows for the easy inclusion of literature citations in our Rmd
# more info here: https://github.com/crsh/citr
# and here:
# http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html
library("citr")
```

```{r load-data, message = FALSE, echo = FALSE}
# load the dataset from a compressed binary file
# it gets loaded as an object called "final_table"
# this has 3.4 million rows...so you will need to be thoughtful about
# how you analyze the data so that you don't overwhelm your laptop
load("output/final_compiled_counts/joined_count_data.RData")

# changes all columns names to lower case so Travis doesn't complain
names(final_table) %<>% tolower

# test that it loaded correctly before proceeding
stopifnot(exists("final_table"))
```

```{r make-summary-table, echo = FALSE}
# There are many many genes in this dataset, so we can
# subset it down to just a few here to look for interesting patterns
# in the most highly expressed
top_15 <- final_table %>%
  group_by(gender, genename) %>%
  summarize(mean_count = mean(counts_lengthscaledtpm)) %>%
  arrange(desc(mean_count)) %>%
  head(n = 15)

# then we can use the `kable()` function to make a nicely formatted
# markdown table
top_15 %>%
  kable()
```

**Table 1**: The most highly expressed genes in both genders included *SFTPB* and *EEF1A1*.

```{r make-barplot-of-highly-expressed-genes, echo = FALSE}
# this code uses the same data as above, but use it to make a
# barplot - remember geom_col() is just like
# when you use geom_bar(stat = "identity")
top_15 %>%
  ggplot(aes(x = genename,
             y = mean_count,
             fill = gender)) +
    geom_col(position = "dodge")
```

**Figure 1**: Here we show an example figure caption.

```{r make-boxplot-of-highly-expressed-genes, echo = FALSE}
# here we just want to pull out the unique gene names and turn
# them into a vector so we can use it below to make a boxplot
# we use the pull() funtion to get this as a vector, just like
# we did when making histograms several weeks ago
top_genes <- top_15 %>%
  ungroup() %>%
  select(genename) %>%
  unique() %>%
  pull()

# now we need to filter from the full data set again, because
# we don't just want summary data, we want all the data in
# order to make boxplots
final_table %>%
  filter(genename %in% top_genes) %>%
  ggplot(aes(x = genename,
             y = counts_lengthscaledtpm,
             fill = gender)) +
    geom_boxplot() +
    facet_wrap(~cancer_stage) +
    xlab("Gene Name") +
    ylab("Scaled read counts per gene") +
    ggtitle("Read counts per gene by gender and cancer stage") +
    theme_bw() + # simplifies theme
    theme(axis.text.x = # rotates x axis labels vertically
            element_text(angle = 90,
                         hjust = 1))
```

**Figure 2**: Here we show another example figure caption.

```{r make-summary-table-SNOR-top15-genes}
#here I want to pull out just the top 15 genes that either 
#have SNORA- or SNORD- in their base name 
#have to filter it out as a subset from final_table
snor_genes <- final_table %>%
  filter(grepl("SNOR", genename)) %>%
  group_by(genename) %>%
  summarize(mean_count = mean(counts_lengthscaledtpm)) %>%
  arrange(desc(mean_count)) %>%
  head(15)

# then we can use the `kable()` function to make a nicely formatted
# markdown table
snor_genes %>%
  kable()
```

 
```{r make-summary-SNOR-table, echo = FALSE}
# There are many many genes in this dataset, so we can
# subset it down to just a few here to look for interesting patterns
# in the most highly expressed
all_snor <- final_table %>%
  filter(grepl("SNOR", genename)) %>%
  group_by(gender, genename, smoking_status, cancer_stage) %>%
  summarize(mean_count = mean(counts_lengthscaledtpm)) %>%
  arrange(desc(mean_count)) %>%
  head(n = 15)

# then we can use the `kable()` function to make a nicely formatted
# markdown table
all_snor %>%
  kable()
```
**Figure 1**:
```{r make-barplot-of-highly-expressed-genes, echo = FALSE}
# this code uses the same data as above, but use it to make a
# barplot - remember geom_col() is just like
# when one would use geom_bar(stat = "identity")
all_snor %>%
  ggplot(aes(x = genename,
             y = mean_count,
             fill = gender)) +
    geom_col(position = "dodge")
```
**Figure 2**:
# Discussion

Add around 1-2 pages interpreting your results and considering future directions one might take in analyzing these data.

# Sources Cited
